---
title: 'Assignment 2: Coding Basics'
author: "Amanda Braun"
geometry: margin=2.54cm
output:
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 


3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. seq is function to generate a sequence, (from, to, by)  why doens't it produce the numbers? 
amanda_sequence <- seq(1, 100, 4) 

#2. 
mean(amanda_sequence)

median(amanda_sequence) 

#3. 
mean(amanda_sequence) < median(amanda_sequence)

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
names <- c("Amanda", "Rani", "Cate", "Rachel") 
names
#character vector 
testscore <- c(95, 48, 92, 72)
testscore
#numeric vector 
testpass <- c(TRUE, FALSE, TRUE, TRUE)
testpass
#logical vector 

classscoresNovember10 <- data.frame(names, testscore, testpass)
classscoresNovember10
names(classscoresNovember10) <- c("Name", "Score", "Pass"); View(classscoresNovember10)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: Both data frames and matrixes represent two dimensional data sets. A data frame can display multiple types of data - character, logical, numeric. A matrix represents one type of data in a set number of rows and columns.  

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
passingscore <- function(x){
  if (x < 50) (x=FALSE) else (x=TRUE)} # if and else function without print 
passingscore (45) 
passingscore(testscore)

passingscoreprint<- function (x) {if (x<50) print("True") else print("False")} # "if" and  "else" function with print 
passingscoreprint(45) #testing single object 
passingscoreprint(testscore) #testing testscore vector 


passing_score2print <- function(x) {
  ifelse((x<50), 
         print("False"), print("True"))} #ifelse function with print 
passing_score2print(6)

passing_score2print(testscore)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: #"if" and "else"  work for one value, but do not work for a vector. When "if" and "else" is used for a vector, an warning message is produced that the conditions has a length > 1 and only the first element will be used, meaning that we cannot input a vector with multiple objects into the function. 
#"ifelse" does work for a vector both a single object and for a vector  
