---
title: "Assignment 9: Spatial Analysis"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on spatial analysis. 

## Directions
1. Use this document to create code for a map. You will **NOT** be turning in the knitted Rmd file this time, only the pdf output for a map.
2. When you have produced your output, submit **only** the pdf file for the map, without any code. Please name your file "StudentName_A09_Spatial.pdf".

The completed exercise is due on Thursday, March 19 at 1:00 pm.

## Create a map

You have three options for this assignment, and you will turn in just **one** final product. Feel free to choose the option that will be most beneficial to you. For all options, to earn full points you should use best practices for data visualization that we have covered in previous assignments (e.g., relabeling axes and legends, choosing non-default color palettes, etc.).

Here are your three options: 

1. Reproduce figure 1b from the spatial lesson, found in section 3.2.2. You may choose a state other than North Carolina, but your map should still contain the spatial features contained in figure 1b in the "img" folder. 

2. Create a new map that mixes spatial and tabular data, as in section 3.3 of the spatial lesson. You may use the maps created in the lesson as an example, but your map should contain data other than precipitation days per year. This map should include: 

* State boundary layer
* Basin boundary layer
* Gage layer
* Tabular data (as an aesthetic for one of the layers)

3. Create a map of any other spatial data. This could be data from the spatial lesson, data from our other course datasets (e.g., the Litter dataset includes latitude and longitude of trap sites), or another dataset of your choosing.  Your map should include: 

* One or more layers with polygon features (e.g., country boundaries, watersheds)
* One or more layers with point and/or line features (e.g., sampling sites, roads)
* Tabular data that correpond to one of the layers, specified as an aesthetic (e.g., total litter biomass at each trap, land cover class at each trap)

Hint: One package that may come in handy here is the `maps` package, which contains several options for basemaps that cover political and geologic boundaries.
```{r}
library("readr")
library("dplyr")
library("tidyr")
library("ggplot2")
library("purrr")
library("sf")
library("ggmap")
library("here")
```

```{r}
pdf(here("outputs", "pdf_test.pdf"), width = 11, height = 8.5)
ggplot(data = cars) +
  geom_point(aes(x = dist, y = speed))
dev.off()
```

```{r}
basins_nf_seplains_raw <- st_read(here("data", "spatial_data", "bas_nonref_SEPlains.shp"))
gages_raw <- st_read(here("data", "spatial_data", "gagesII_9322_sept30_2011.shp"))
southeast_state_bounds_raw <- st_read(here("data", "spatial_data", "southeast_state_bounds.shp"))
my_tabular_data_raw <- read_csv(here("data", "tabular_data", "conterm_climate.txt")) 
```

```{r}
my_proj4 <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
my_epsg <- 5070

basins_nf_seplains <- basins_nf_seplains_raw
st_crs(basins_nf_seplains) <- my_proj4
basins_nf_seplains <- basins_nf_seplains %>%
  st_set_crs(my_epsg)
st_crs(basins_nf_seplains)


gages <- gages_raw
st_crs(gages) <- my_proj4
gages <- gages %>%
  st_set_crs(my_epsg)
st_crs(gages)

southeast_state_bounds <- southeast_state_bounds_raw
st_crs(southeast_state_bounds) <- my_proj4
southeast_state_bounds <- southeast_state_bounds %>%
  st_set_crs(my_epsg)
st_crs(southeast_state_bounds)


na_albers_proj4 <- "+proj=aea +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
na_albers_epsg <- 102008

southeast_state_bounds_na_albers <-  sf::st_transform(southeast_state_bounds, crs = na_albers_proj4) %>%
  st_set_crs(na_albers_epsg)
```


```{r}
```


```{r}
st_crs(basins_nf_seplains)
st_crs(gages)
st_crs(southeast_state_bounds)
st_crs(southeast_state_bounds_na_albers)
```

```{r}
#select  North Carolina (NC)
nc_state_bounds_geom <- southeast_state_bounds %>%
  filter(NAME == "North Carolina") %>%
  st_geometry()

# select watersheds that intersect with NC bounds
nc_basins_nf_seplains <- basins_nf_seplains %>%
  st_intersection(nc_state_bounds_geom)

# check
head(nc_basins_nf_seplains)

# select gages that fall within NC bounds
# add your code here
nc_gages <- gages %>%
  st_intersection(nc_state_bounds_geom)

head(nc_gages)
names(nc_gages)

names(my_tabular_data_raw)

names(nc_gages)
# use "STAID"

# join the tabular data to nc_gages
nc_gages$STAID <- as.factor(nc_gages$STAID)
  left_join(my_tabular_data_raw, by = "STAID")

my_tabular_data_raw$STAID <-as.factor(my_tabular_data_raw$STAID)
  
# check that it worked
names(nc_gages_climate)

nc_gages_climate <- nc_gages %>%
  left_join(my_tabular_data_raw, by = "STAID")



```
```{r}

```
```{r}
# select North Carolina (NC)
nc_state_bounds_geom <- southeast_state_bounds %>%
  filter(NAME == "North Carolina") %>%
  st_geometry()
##dropping geometry 

# get data bounding box
##buffer north carolina boundary by 1500000 meters can use trial and errr to figure out what is big enough 
nc_bbox <- nc_state_bounds_geom %>%
  st_buffer(dist = 150000) %>% 
  st_transform(4326) %>% 
  st_bbox()
nc_bbox 

# fix bounding box columns so they match what is needed for the 

nc_bbox_fix <- c(left = nc_bbox[[1]], bottom = nc_bbox[[2]], right = nc_bbox[[3]], top = nc_bbox[[4]])

# check result
nc_bbox_fix

# get basemap
nc_basemap <- get_map(nc_bbox_fix, maptype = 'terrain-background', source = 'stamen', zoom = 8)
##loads tiles 
class(nc_basemap)
##raster
st_crs(nc_basemap)

##stamen basemap, other people use different type of basemaps
## help -> get_map, tells you the different zoom scales 
# you'll have to trust me that this will have a CRS = WGS84

# convert nc_state_bounds_geom to WGS84 so it matches nc_basemap
nc_state_bounds_geom_wsg84 <- nc_state_bounds_geom %>%
  st_transform(4326)
##project back to longitutde and latitude 

# check
st_crs(nc_state_bounds_geom_wsg84)

# plot without basemap
pdf(here("outputs", "nc_without_basemap.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = nc_state_bounds_geom_wsg84, fill = NA, lwd = 2) +
  theme_bw()
dev.off()

# plot with basemap
# windows version?
pdf(here("outputs", "nc_with_basemap.pdf"), width = 11, height = 8.5)
ggmap(nc_basemap) +
  geom_sf(data = nc_state_bounds_geom_wsg84, fill = NA, lwd = 1, inherit.aes = FALSE)
```

```{r}
pdf(here("outputs", "Spatial_Data_AssignmentX.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = nc_state_bounds_geom_wsg84, fill = NA, lwd = 1, inherit.aes = FALSE) +
  geom_sf(data = nc_basins_nf_seplains, alpha = 0.25, color = "gray") +
  geom_sf(data = nc_gages, lwd = 1) +
  geom_sf(data = nc_gages_climate, aes(color = T_AVG_SITE), size = 1) +
  scale_color_gradient(low = "white", high = "blue", labels = c("10° C", "12° C", "14° C ", "16° C")) +
  labs(color = "Stream Gage Average Annual Air Temp °C )") +
  theme_bw()
dev.off()


```

```{r}
sc_state_bounds_geom <- southeast_state_bounds %>%
  filter(NAME == "South Carolina") %>%
  st_geometry()

# select watersheds that intersect with NC bounds
sc_basins_nf_seplains <- basins_nf_seplains %>%
  st_intersection(sc_state_bounds_geom)

# check
head(sc_basins_nf_seplains)

# select gages that fall within NC bounds
# add your code here
sc_gages <- gages %>%
  st_intersection(sc_state_bounds_geom)

head(sc_gages)
names(sc_gages)

names(my_tabular_data_raw)

names(sc_gages)
# use "STAID"

# join the tabular data to nc_gages
sc_gages$STAID <- as.factor(sc_gages$STAID)
  left_join(my_tabular_data_raw, by = "STAID")

my_tabular_data_raw$STAID <-as.factor(my_tabular_data_raw$STAID)
  
# check that it worked


sc_gages_climate <- sc_gages %>%
  left_join(my_tabular_data_raw, by = "STAID")

names(sc_gages_climate)


```
```{r}

```
```{r}
# select North Carolina (NC)
sc_state_bounds_geom <- southeast_state_bounds %>%
  filter(NAME == "South Carolina") %>%
  st_geometry()
##dropping geometry 

# get data bounding box
##buffer north carolina boundary by 1500000 meters can use trial and errr to figure out what is big enough 
sc_bbox <- sc_state_bounds_geom %>%
  st_buffer(dist = 150000) %>% 
  st_transform(4326) %>% 
  st_bbox()
sc_bbox 

# fix bounding box columns so they match what is needed for the 

sc_bbox_fix <- c(left = sc_bbox[[1]], bottom = sc_bbox[[2]], right = sc_bbox[[3]], top = sc_bbox[[4]])

# check result
sc_bbox_fix

# get basemap
sc_basemap <- get_map(sc_bbox_fix, maptype = 'terrain-background', source = 'stamen', zoom = 8)
##loads tiles 
class(sc_basemap)
##raster
st_crs(sc_basemap)

##stamen basemap, other people use different type of basemaps
## help -> get_map, tells you the different zoom scales 
# you'll have to trust me that this will have a CRS = WGS84

# convert nc_state_bounds_geom to WGS84 so it matches nc_basemap
sc_state_bounds_geom_wsg84 <- sc_state_bounds_geom %>%
  st_transform(4326)
##project back to longitutde and latitude 

# check
st_crs(sc_state_bounds_geom_wsg84)

# plot without basemap
pdf(here("outputs", "sc_without_basemap.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = sc_state_bounds_geom_wsg84, fill = NA, lwd = 2) +
  theme_bw()
dev.off()

# plot with basemap
# windows version?
pdf(here("outputs", "sc_with_basemap.pdf"), width = 11, height = 8.5)
ggmap(sc_basemap) +
  geom_sf(data = sc_state_bounds_geom_wsg84, fill = NA, lwd = 1, inherit.aes = FALSE)+
theme_bw()
dev.off()
```

```{r}
pdf(here("outputs", "AmandaBraun_A09_Spatial.pdf"), width = 11, height = 8.5)
ggplot() +
  geom_sf(data = sc_state_bounds_geom, fill = "white") +
  geom_sf(data = sc_basins_nf_seplains, alpha = 0.25, color = "dark gray") +
  geom_sf(data = sc_gages, lwd = 1) +
  geom_sf(data = sc_gages_climate, aes(color = T_AVG_SITE), size = 1) +
  scale_color_gradient(low = "white", high = "blue", labels = c("14° C", "15° C", "16° C ", "17° C", "18° C")) +
  labs(color = "Stream Gage\nAverage Annual Air Temp °C") +
  theme_bw()
dev.off()


```

```{r}
pdf(here("outputs", "AmandaBraun_A09_Spatial.pdf"), width = 11, height = 8.5)
ggmap(sc_basemap) +
  geom_sf(data = sc_state_bounds_geom_wsg84, fill = NA, lwd = 1, inherit.aes = FALSE)+
  geom_sf(data = sc_basins_nf_seplains, alpha = 0.25, color = "dark gray") +
  geom_sf(data = sc_gages, lwd = 1) +
  geom_sf(data = sc_gages_climate, aes(color = T_AVG_SITE), size = 1) +
  scale_color_gradient(low = "white", high = "blue", labels = c("14° C", "15° C", "16° C ", "17° C", "18° C")) +
  labs(color = "Stream Gage\nAverage Annual Air Temp °C") +
  theme_bw()
dev.off()
```

